---
title: "Expploring disease activation functions"
output: html_notebook
---

In this notebook, we'll look at how changing the disease activation fucntion affects prevalence, heritability, etc.

```{r}
library(tidyverse)
library(MASS)
library(fastmatrix)
library(tidyverse)
library(matlib)

simulate_genotypes <- function(N,L){
  ### function to simulate genotypes
  ps <- rbeta(n = L,4,4)
  G <- matrix(NA,nrow=N,ncol=L)
  for (i in c(1:L)){
    gen <- rbinom(N, size = 2, prob = ps[i])
    G[,i] <- gen
  }
  return(G)
}

give_disease_risk <- function(nodes,f=mean,threshold=5){
  # returns disease risk/severity 
  # risk given by some function of the nodes
  risk <- f(nodes)
  return(risk)
}

# a more funk disease function
mdd_risk <- function(v,threshold=1,n=3){
  # linear combination of node activation - this is just an example
  return(sum(v > threshold)>n)
}

greml <- function( y, G, constrain=F, quantnorm=F ){
  if( quantnorm ){
    ranks <- rank(y)
    quantiles <- qnorm(ranks/(length(ranks) + 1))
    y <- quantiles
  }
  y <- scale(y)
  G <- scale(G)
  N <- nrow(G)
  svdG <- svd(G)
  out <- eigen_lmm( yprime=t(svdG$u) %*% y, Lam.K=(svdG$d/sqrt(N)+1e-12)^2, constrained=constrain )
  sig2g <- out$sig2g
  sig2e <- out$sig2e
  list( h2=sig2g/(sig2g+sig2e), sig2g=sig2g, sig2e=sig2e )
}


eigen_lmm <- function (yprime, Lam.K, constrained = TRUE) {
  if (constrained) {
    delta <-
      optimise(
        eigen_lmm_obj,
        c(0, 1e5),
        maximum = TRUE,
        Lam.K = Lam.K,
        yprime = yprime
      )$maximum
  } else {
    ### these parameters correspond to h^2>0, with h^2>1 for delta<0
    opt_lims_plus <- c(-min(Lam.K), 1e5)
    opt_plus <-
      optimise(
        eigen_lmm_obj,
        opt_lims_plus,
        maximum = TRUE,
        Lam.K = Lam.K,
        yprime = yprime
      )
    ### these parameters correspond to h^2<0
    opt_lims_minus <- c(-1000, -max(Lam.K))
    opt_minus <-
      optimise(
        eigen_lmm_obj,
        opt_lims_minus,
        maximum = TRUE,
        Lam.K = Lam.K,
        yprime = yprime
      )
    delta <-
      ifelse(opt_plus$objective > opt_minus$objective,
             opt_plus$maximum,
             opt_minus$maximum)
  }
  sig2g <- mean(yprime ^ 2 / (Lam.K + delta))
  sig2e <- delta * sig2g
  list(h2 = sig2g / (sig2g + sig2e),
       sig2g = sig2g,
       sig2e = sig2e)
}


eigen_lmm_obj <- function(delta, yprime, Lam.K){
  denominator <- Lam.K + delta
  sig2g <- mean(yprime^2/denominator)
  sum(dnorm( yprime, mean=0, sd=sqrt(sig2g*denominator), log=TRUE ))}

sig_from_row <- function(ind.sig,i,n_node){
  ## function which takes the matrix of individ. level sigmas and builds a covariance matrix which we can draw from
  
  # extract relevant row
  sig_vec <- ind.sig[i,]
  
  # make into matrix
  sig_mat <- matrix(0,n_node,n_node) 
  sig_mat[lower.tri(sig_mat,diag=TRUE)] <- sig_vec
  sig_mat[upper.tri(sig_mat)] <- t(sig_mat)[upper.tri(sig_mat)]
  
  # how to build diagonal? Here I will add the environmental variances
  #diag(sig_mat) <- diag(S$Sigma) + diag(sig_mat)
  
  return(sig_mat)
}


theme_set(theme_bw())
```

```{r}
l <- 200 # number of loci
N <- 1000 # number of individuals
P <- 1 # number of symptoms 
h2 <- 0.9
reps <- 10
heritabilities <- seq(0,1,0.2)
```


## Null model: MDD is a function of symptoms, symptoms are IID

```{r}
G <- simulate_genotypes(N = N,L = l)
G <- scale(G)
beta <- matrix(rnorm(P * l),nrow=l) %*% diag(1,P,P) * sqrt(h2 / l)
e <- matrix(rnorm(P * N),nrow=N) * sqrt(1-h2)

X <- G %*% beta + e
cov(e)
cov(G %*% beta)
cov(X)
greml(X[,1],G)

Y <- apply(X,1,mdd_risk,threshold=0)
greml(Y,G)

plot(rowSums(X),Y)
```
```{r}
rows <- data.frame()


for (r in c(1:reps)){
  print(r)
  for (her in heritabilities){
    
    G <- simulate_genotypes(N = N,L = l)
    G <- scale(G)
    beta <- matrix(rnorm(P * l),nrow=l) %*% diag(1,P,P) * sqrt(her / l)
    e <- matrix(rnorm(P * N),nrow=N) * sqrt(1-her)
  
    X <- G %*% beta + e
    Y <- apply(X,1,mdd_risk,threshold=0.5,n=0)
    
    res <- greml(Y,G)
    h2 <- res$h2
    
    rows <- rbind(rows,data.frame("h2_symptoms"=her,
                                  "h2_disease"=h2,
                                  "sig2g"=res$sig2g,
                                  "sig2e"=res$sig2e,
                                  "prev"=sum(Y),
                                  "rep"=r))
    
  }
}

rows
```
```{r}
rows <- data.frame()
reps <- 10
nsymp <- 5

for (r in c(1:reps)){
  print(r)
  for (her in heritabilities){
    for (p in c(1:nsymp)){
    
      G <- simulate_genotypes(N = N,L = l)
      G <- scale(G)
      beta <- matrix(rnorm(p * l),nrow=l) %*% diag(1,p,p) * sqrt(her / l)
      e <- matrix(rnorm(p * N),nrow=N) * sqrt(1-her)
    
      X <- G %*% beta + e
      
      c <- 1
      Y <- apply(X,1,mdd_risk,threshold=c,n=p-1)
      prev <- sum(Y)/N
      
      while (prev<0.25){
        c <- c - 0.1
        Y <- apply(X,1,mdd_risk,threshold=c,n=p-1)
        prev <- sum(Y)/N
      }
      
      res <- greml(Y,G)
      h2 <- res$h2
      
      rows <- rbind(rows,data.frame("h2_symptoms"=her,
                                    "h2_disease"=h2,
                                    "sig2g"=res$sig2g,
                                    "sig2e"=res$sig2e,
                                    "prev"=sum(Y),
                                    "rep"=r,
                                    "P"=p,
                                    "c"=c))
    }
  }
}

rows
```

```{r}
rows <- rows %>% mutate(h2_l = h2_disease * (prev/N) * (1-prev/N) / dnorm(prev/N)^2)
```

## How is disease heritability affected by the number of symptoms constituting the disease?

```{r}
ggplot(rows,aes(x=h2_symptoms,y=h2_disease)) +
  geom_point(aes(col="oberved h2 of disease (GREML)")) +
  geom_point(aes(y=h2_l,col="expected h2_l (one symptom)")) +
  geom_abline(slope=1,intercept=0,lty=2,col="darkgrey") +
  facet_wrap(~P,nrow=1) +
  labs(title = "Number of symptoms constituting disease",
     col="",
     subtitle = "Independent Symptoms")
```

```{r}
rows_n <- data.frame()
reps <- 10
p <- 5

for (r in c(1:reps)){
  print(r)
  for (her in heritabilities){
    for (n in c(0:(p-1))){
      G <- simulate_genotypes(N = N,L = l)
      G <- scale(G)
      beta <- matrix(rnorm(p * l),nrow=l) %*% diag(1,p,p) * sqrt(her / l)
      e <- matrix(rnorm(p * N),nrow=N) * sqrt(1-her)
    
      X <- G %*% beta + e
      
      c <- 1
      Y <- apply(X,1,mdd_risk,threshold=c,n=n)
      prev <- sum(Y)/N
      
      while (prev<0.25){
        c <- c - 0.1
        Y <- apply(X,1,mdd_risk,threshold=c,n=n)
        prev <- sum(Y)/N
      }
      
      res <- greml(Y,G)
      h2 <- res$h2
      
      rows_n <- rbind(rows_n,data.frame("h2_symptoms"=her,
                                    "h2_disease"=h2,
                                    "sig2g"=res$sig2g,
                                    "sig2e"=res$sig2e,
                                    "prev"=sum(Y),
                                    "rep"=r,
                                    "P"=p,
                                    "c"=c,
                                    "n"=n))
    }
  }
}

rows_n

rows_n <- rows_n %>% mutate(h2_l = h2_disease * (prev/N) * (1-prev/N) / dnorm(prev/N)^2)

```
## How is disease heritability affected by the number of symptoms required past a threshold for the disease - for a given number of symptoms?


```{r}
ggplot(rows_n,aes(x=h2_symptoms,y=h2_disease)) +
  geom_point(aes(col="oberved h2 of disease (GREML)")) +
  geom_point(aes(y=h2_l,col="expected h2_l (one symptom)")) +
  geom_abline(slope=1,intercept=0,lty=2,col="darkgrey") +
  facet_wrap(~(n+1),nrow=1) +
  labs(title = paste("Number of symptoms required past threshold out of",p),
       col="",
       subtitle = "Independent Symptoms")
```
```{r}
ggplot(rows_n,aes(x=h2_symptoms,y=h2_disease)) +
  geom_point(aes(col=factor(n)) )
```



```{r}
ggplot(rows_n,aes(x=h2_symptoms,y=prev/N,col=factor(n))) +
  geom_point(aes(shape="prevalence")) +
  labs(y="disease prevalence") +
  geom_point(aes(y=c,shape="cut-off")) +
  lims(y=c(0,1))
```

```{r}
ggplot(rows,aes(x=h2_symptoms,y=sig2g)) +
  geom_point(aes(col="sig2g") )+
  geom_point(aes(y=sig2e,col="sig2e"))
```

## Slightly less null model: MDD is a function of symptoms, symptoms have a fixed covariance structure

```{r}
# generate a random covariance matrix for symptoms
c <- 0.3
Sigma <- matrix(sample(x=c(0,c),size=P * P,replace=TRUE),byrow = T,ncol=P) # draw covariances 
diag(Sigma) <- 1 # set diagonal
Sigma[lower.tri(Sigma)] = t(Sigma)[lower.tri(Sigma)] # make symmetric

# print av of correlations
```



```{r,eval=FALSE}
rows_2 <- data.frame()

for (r in c(1:reps)){
  for (her in heritabilities){
    G <- simulate_genotypes(N = N,L = l)
    G <- scale(G)
    beta <- matrix(mvrnorm(n=l,mu = rep(0,P),Sigma = Sigma),nrow=l) %*% diag(1,P,P) * sqrt(her / l)
    e <- matrix(mvrnorm(n=N,mu = rep(0,P),Sigma = Sigma),nrow=N) * sqrt(1-her)
  
    X <- G %*% beta + e
    Y <- apply(X,1,mdd_risk,threshold=0)
    
    res <- greml(Y,G)
    h2 <- res$h2
    
    rows_2 <- rbind(rows_2,data.frame("h2_symptoms"=her,
                                  "h2_disease"=h2,
                                  "sig2g"=res$sig2g,
                                  "sig2e"=res$sig2e,
                                  "prev"=sum(Y),
                                  "rep"=r))
    
  }
}

rows_2
```

```{r,eval=F}
ggplot(rows,aes(x=h2_symptoms,y=h2_disease)) +
  geom_point() +
  geom_point(data=rows_2,col="orchid",shape=1)
```


## Our model: MDD is a function of symptoms, covariance structure variable among individuals

Genetic effects on entries of covariance matrix directly.


```{r,eval=FALSE}
rows_3 <- data.frame()

num_covariances <- (P^2-P)/2 +P

for (r in c(1:reps)){
  for (her in heritabilities){
    for (hercov in cov_heritabilities){
      G <- simulate_genotypes(N = N,L = l)
      G <- scale(G)
      
      
      beta_cov <- matrix(rnorm(num_covariances * l),nrow=l) %*% diag(1,num_covariances,num_covariances) * sqrt(hercov / l)
      e_cov <- matrix(rnorm(N * num_covariances),nrow=N) * sqrt(1-hercov)
      covs <- G %*% beta_cov + e_cov
      
      beta <- matrix(rnorm(P * l),nrow=l) %*% diag(1,P,P) * sqrt(her / l)
      e <- matrix(rnorm(P * N),nrow=N) * sqrt(1-her)
      
      X <- matrix(0,nrow=N,ncol=P)
      
      sig_gg_vecs <- G %*% beta_cov
      #sig_from_row(sig_gg_vecs,1,P) / 100 + Sigma
      for (ind in c(1:N)){
          # make into matrix
          sig_gg_mat <- matrix(0,P,P) 
          sig_gg_mat[lower.tri(sig_gg_mat,diag=FALSE)] <- sig_gg_vec
          sig_gg_mat[upper.tri(sig_gg_mat)] <- t(sig_gg_mat)[upper.tri(sig_gg_mat)]
          
          sig_ee_vec <- e_cov[ind,]
          sig_ee_mat <- matrix(0,P,P) 
          sig_ee_mat[lower.tri(sig_ee_mat,diag=FALSE)] <- sig_ee_vec
          sig_ee_mat[upper.tri(sig_ee_mat)] <- t(sig_ee_mat)[upper.tri(sig_ee_mat)]
  
          
  
          Sig_ind <- Sigma + sig_gg_mat/100 + sig_ee_mat/100 # change scaling constant?
          # print(sum(eigen(Sig_ind)$values<0)==0)
          beta_transformed <- beta %*% t(chol(Sig_ind)) # check this
          e_transformed <- e[ind,] %*% t(chol(Sig_ind)) # check this
          X[ind,] <- G[ind,] %*% beta_transformed + e_transformed
          print(Sig_ind)
      }
      
      Y <- apply(X,1,mdd_risk,threshold=0)
      
      res <- greml(Y,G)
      h2 <- res$h2
      
      rows_3 <- rbind(rows_3,data.frame("h2_symptoms"=her,
                                    "h2_disease"=h2,
                                    "h2_covariance"=hercov,
                                    "sig2g"=res$sig2g,
                                    "sig2e"=res$sig2e,
                                    "prev"=sum(Y),
                                    "rep"=r))
      
    }
  }
}

rows_3
```

```{r}
rows_3 <- data.frame()
cov_heritabilities <- seq(0,1,0.2)
num_covariances <- (P^2-P)/2 +P
reps <- 5

for (r in c(1:reps)){
  for (her in heritabilities){
    for (hercov in cov_heritabilities){
      print(paste(r,her,hercov))
      G <- simulate_genotypes(N = N,L = l)
      G <- scale(G)
      
      
      beta_cov <- matrix(rnorm(num_covariances * l),nrow=l) %*%
        diag(1,num_covariances,num_covariances) * sqrt(hercov / l)
      e_cov <- matrix(rnorm(N * num_covariances),nrow=N) * sqrt(1-hercov)
      covs <- G %*% beta_cov + e_cov
      
      beta <- matrix(rnorm(P * l),nrow=l) %*% diag(1,P,P) * sqrt(her / l)
      e <- matrix(rnorm(P * N),nrow=N) * sqrt(1-her)
      
      X <- matrix(0,nrow=N,ncol=P)
      
      sig_gg_vecs <- G %*% beta_cov
      
      for (ind in c(1:N)){
          # make into matrix
          Sig_ind <- sig_from_row(sig_gg_vecs,ind,P) / 20 + sig_from_row(e_cov,ind,P) / 20 +  Sigma
          eigs <- eigen(Sig_ind)
          eig_vals <- eigs$values
          eig_vecs <- eigs$vectors
          if (sum(eig_vals < 0) > 0) {
              print("Invalid covariance matrix")
              eig_vals[eig_vals < 0] <- 0.01  # Replace negative eigenvalues with a small positive value
              Sig_ind <- eig_vecs %*% diag(eig_vals) %*% t(eig_vecs)
          }
          # 
          beta_transformed <- beta %*% t(chol(Sig_ind)) # check this
          e_transformed <- e[ind,] %*% t(chol(Sig_ind)) # check this
          X[ind,] <- G[ind,] %*% beta_transformed + e_transformed
      }
      
      Y <- apply(X,1,mdd_risk,threshold=0)
      
      res <- greml(Y,G)
      h2 <- res$h2
      
      rows_3 <- rbind(rows_3,data.frame("h2_symptoms"=her,
                                    "h2_disease"=h2,
                                    "h2_covariance"=hercov,
                                    "sig2g"=res$sig2g,
                                    "sig2e"=res$sig2e,
                                    "prev"=sum(Y),
                                    "rep"=r))
      
    }
  }
}

rows_3
```



```{r}
ggplot(rows_3,aes(x=h2_symptoms,y=h2_disease,col=factor(h2_covariance))) +
  geom_point() +
  geom_smooth(se=F) +
  labs(col="covariance heritability")
```


## Symptoms are not heritable and there is covariance

```{r}
rows_n <- data.frame()
reps <- 10
p <- 5
rho <- 0.5
her <- 0

  
for (r in c(1:reps)){
  print(r)
  for (rho in seq(0,1,0.2)){
    for (n in c(1:(p-1))){
      print(paste("Running for rep",r, "heritability", her, "threshold number", n, "rho", rho))
      G <- simulate_genotypes(N = N,L = l)
      G <- scale(G)
      
      Sigma <- matrix(rho,nrow=p,ncol=p)
      diag(Sigma) <- 1 
      
      beta <- matrix(mvrnorm(n=l,mu = rep(0,p),Sigma = Sigma),nrow=l) %*% diag(1,p,p) * sqrt(her / l)
      e <- matrix(mvrnorm(n=N,mu = rep(0,p),Sigma = Sigma),nrow=N) * sqrt(1-her)
    
      X <- G %*% beta + e
      
      c <- 2
      Y <- apply(X,1,mdd_risk,threshold=c,n=n)
      prev <- sum(Y)/N
      
      while (prev<0.25){
        c <- c - 0.1
        Y <- apply(X,1,mdd_risk,threshold=c,n=n)
        prev <- sum(Y)/N
      }
      
      print(paste("Set threshold",c,"and prevalence",prev))
      res <- greml(Y,G)
      h2 <- res$h2
      
      rows_n <- rbind(rows_n,data.frame("h2_symptoms"=her,
                                    "h2_disease"=h2,
                                    "sig2g"=res$sig2g,
                                    "sig2e"=res$sig2e,
                                    "prev"=sum(Y),
                                    "rep"=r,
                                    "P"=p,
                                    "c"=c,
                                    "n"=n,
                                    "rho"=rho))
    }
  }
}

rows_n

rows_n <- rows_n %>% mutate(h2_l = h2_disease * (prev/N) * (1-prev/N) / dnorm(prev/N)^2)

```
```{r}
ggplot(rows_n,aes(x=n+1,y=h2_disease)) +
  geom_point(aes(col=factor(rho))) +
  #geom_point(aes(y=h2_l,col="expected h2_l (one symptom)")) +
  labs(title = paste("Number of symptoms required past threshold out of",p),
       col="",
       subtitle = "Independent Symptoms")
```

```{r}
ggplot(rows_n,aes(x=n+1,y=c)) +
  geom_jitter(aes(col=factor(rho)),height = 0,width=0.2) +
  #geom_point(aes(y=h2_l,col="expected h2_l (one symptom)")) +
  labs(title = paste("Number of symptoms required past threshold out of",p),
       col="Covariance parameter",
       x="Number of symptoms past threshold",
       y="threshold")
```

## Symptoms are not heritable, but covariance is heritable

We model genetic effects on covariance through the diagonal matrix $D$ in the $LDL^T$ decomposition of $\Sigma$. This works in the following way:

$D$ is an $(N \times P)$ matrix of diagonal entries. Each entry of D is independent of the others, but individuals co-vary in accordance with their kinship:


$$
D \sim MN(\mathbf{M},\vec{\sigma^2_{cov,e}}\mathbf{I}_n+\vec{\sigma^2_{cov,g}}\mathbf{K},\mathbf{I}_P)
$$
where $\mathbf{M}$ is an $(N \times P)$ matrix where each row is identical, and is the vector $\vec{\sigma^2_X}$. Between rows, there is independent environmental noise as well as genetic effects.




$$
D = G\beta_{cov} + \epsilon
$$
The effect size matrix is $(P \times l)$. $\epsilon_i \sim N(\vec{0}^{p},\vec{\sigma^2_{cov,e}})$. Since everyone has slightly different $D$, each individual's personal symptom covariance matrix varies: 

$$
\Sigma_i = LD_iL^T
$$
And from this, the symptoms are drawn:

$$
X_i \sim N(\vec{0}^p, \Sigma_i)
$$

```{r}
rows_hercov <- data.frame()
reps <- 10
p <- 5
her <- 0

L <- matrix(0,nrow=p,ncol=p)
L[lower.tri(L,diag=T)] <- 1
  
for (r in c(1:reps)){
  print(r)
  for (hercov in seq(0,1,0.2)){
    for (n in c(1:(p-1))){
      print(paste("Running for rep",r, "heritability", her, "threshold number", n, "covariance heritability", hercov))
      G <- simulate_genotypes(N = N,L = l)
      G <- scale(G)
      
      # effect sizes on D
      beta_cov <- matrix(mvrnorm(n=l,mu = rep(0,p),Sigma = diag(p)),nrow=l) %*% diag(3,p,p) * sqrt(hercov / l)
      e_cov <- matrix(mvrnorm(n=N,mu = rep(0,p),Sigma = diag(p)),nrow=N) %*% diag(3,p,p) * sqrt(1-hercov)
      # individual-level D
      D_vectors <- G %*% beta_cov + e_cov + matrix(10,nrow=N,ncol=p)
      D_vectors[D_vectors<0] <- 0.001
      
      # effect sizes on trait directly
      beta <- matrix(mvrnorm(n=l,mu = rep(0,p),Sigma = diag(5,p,p)),nrow=l) %*% diag(1,p,p) * sqrt(her / l)
      e <- matrix(mvrnorm(n=N,mu = rep(0,p),Sigma = diag(5,p,p)),nrow=N) * sqrt(1-her)
      
      X <- matrix(0,nrow=N,ncol=p)
      # Give each person a covariance matrix and draw symptoms
      for (ind in c(1:N)){
          # make into matrix
          Sig_ind <- L %*% diag(D_vectors[ind,]) %*% t(L)
          beta_transformed <- t(chol(Sig_ind)) %*% t(beta) %>% t() # check this
          e_transformed <- t(chol(Sig_ind)) %*% e[ind,] %>% t() # check this
          X[ind,] <- G[ind,] %*% beta_transformed + e_transformed
      }
      
      # setting the threshold to keep constant prevalence
      c <- 10
      Y <- apply(X,1,mdd_risk,threshold=c,n=n)
      prev <- sum(Y)/N
      
      while (prev<0.25){
        c <- c - 0.1
        Y <- apply(X,1,mdd_risk,threshold=c,n=n)
        prev <- sum(Y)/N
      }
      
      print(paste("Set threshold",c,"and prevalence",prev))
      res <- greml(Y,G)
      h2 <- res$h2
      
      rows_hercov <- rbind(rows_hercov,data.frame("h2_symptoms"=her,
                                    "h2_disease"=h2,
                                    "sig2g"=res$sig2g,
                                    "sig2e"=res$sig2e,
                                    "prev"=sum(Y),
                                    "rep"=r,
                                    "P"=p,
                                    "c"=c,
                                    "n"=n,
                                    "hercov"=hercov))
    }
  }
}

rows_hercov

rows_hercov <- rows_hercov %>% mutate(h2_l = h2_disease * (prev/N) * (1-prev/N) / dnorm(prev/N)^2)

```

```{r}
ggplot(rows_hercov,aes(x=hercov,y=h2_disease)) +
  geom_point(aes(col=factor(n+1))) +
  #geom_point(aes(y=h2_l,col="expected h2_l (one symptom)")) +
  labs(x="Covariance heritability",
       col="Number of symptoms past threshold",
       y="Disease heritability")
```

```{r}
ggplot(rows_hercov,aes(x=hercov,y=h2_l)) +
  geom_point(aes(col=factor(n+1))) 
```

## GWAS on individual-level products


```{r}
N <- 5000
l <- 200
G <- simulate_genotypes(N = N,L = l)
G <- scale(G)
p <- 5
hercov <- .9
her <- .1
L <- matrix(0,nrow=p,ncol=p)
L[lower.tri(L,diag=T)] <- 1
diag(L) <- 1

beta_cov <- matrix(mvrnorm(n=l,mu = rep(0,p),Sigma = diag(p)),nrow=l) %*% diag(3,p,p) * sqrt(hercov / l)
e_cov <- matrix(mvrnorm(n=N,mu = rep(0,p),Sigma = diag(p)),nrow=N) %*% diag(3,p,p) * sqrt(1-hercov)
D_vectors <- G %*% beta_cov + e_cov + matrix(10,nrow=N,ncol=p)
D_vectors[D_vectors<0] <- 0.1


beta <- matrix(mvrnorm(n=l,mu = rep(0,p),Sigma = diag(5,p,p)),nrow=l) %*% diag(1,p,p) * sqrt(her / l)
e <- matrix(mvrnorm(n=N,mu = rep(0,p),Sigma = diag(5,p,p)),nrow=N) * sqrt(1-her)

X <- matrix(0,nrow=N,ncol=p)
# Give each person a covariance matrix and draw symptoms
for (ind in c(1:N)){
    # make into matrix
    Sig_ind <- L %*% diag(D_vectors[ind,]) %*% t(L)
    beta_transformed <- t(chol(Sig_ind)) %*% t(beta) %>% t() # check this
    e_transformed <- t(chol(Sig_ind)) %*% e[ind,] %>% t() # check this
    X[ind,] <- G[ind,] %*% beta_transformed + e_transformed
}

# setting the threshold to keep constant prevalence
c <- 1
n <- 3
Y <- apply(X,1,mdd_risk,threshold=c,n=n)
prev <- sum(Y)/N
```


```{r}
lm(X.1 *X.2 ~ D.1 + D.2 + D.3, data= data.frame("X"=X,"D"=D_vectors)) %>% summary()
lm(X.1 > c & X.2 > c ~ D.1, data= data.frame("X"=X,"D"=D_vectors)) %>% summary()

data.frame("X"=X,"D"=D_vectors,"Y"=Y) %>%
ggplot(aes(y=X.1*X.2,x=D.1)) + geom_point() +
  geom_smooth()

data.frame("X"=X,"D"=D_vectors,"Y"=Y) %>%
ggplot(aes(y=X.1*X.2,x=Y)) + geom_boxplot()


data.frame("X"=X,"D"=D_vectors) %>%
ggplot(aes(y=X.1>c & X.2 > c,x=D.1)) + geom_point() 


data.frame("X"=X,"D"=D_vectors,"Y"=Y) %>% group_by(X.1>c & X.2 > c) %>% summarise(D = mean(D.1),X12 = mean(X.1*X.2),Yprv= mean(Y))
```


```{r}
IP <- X[,1]*X[,2]

res <- apply(G , 2, function(x) lm(IP ~ x))
p_hats <- sapply(res, function(x) summary(x)$coefficients[2, 4])
beta_hats <- sapply(res, function(x) summary(x)$coefficients[2, 1])


plot(beta_cov[,1],beta_hats)

lm(beta_hats ~ beta_cov[,1]) %>% summary()
summary(res[[1]])$coefficients[2,1]
```


```{r}
cov(X[,1:5])/5
colMeans(D_vectors)
```


```{r}
rows_hercov2 <- data.frame()
reps <- 5
p <- 5
her <- 0

D <- diag(1,p,p)
L[lower.tri(L,diag=T)] <- 1
  
for (r in c(1:reps)){
  print(r)
  for (hercov in seq(0,1,0.2)){
    for (n in c(1:(p-1))){
      print(paste("Running for rep",r, "heritability", her, "threshold number", n, "covariance heritability", hercov))
      G <- simulate_genotypes(N = N,L = l)
      G <- scale(G)
      K <- G %*% t(G) / l
      
      n_l <- (p ^ 2 - p) / 2
      beta_l <- matrix(mvrnorm(n=l,mu = rep(0,n_l),Sigma = diag(n_l)),nrow=l) %*% diag(1,n_l,n_l) * sqrt(hercov / l)
      e_l <- matrix(mvrnorm(n=N,mu = rep(0,n_l),Sigma = diag(n_l)),nrow=N) * sqrt(1-hercov)
      L_vectors <- G %*% beta_l + e_l 
      
      
      beta <- matrix(mvrnorm(n=l,mu = rep(0,p),Sigma = D),nrow=l) %*% diag(1,p,p) * sqrt(her / l)
      e <- matrix(mvrnorm(n=N,mu = rep(0,p),Sigma = D),nrow=N) * sqrt(1-her)

      for (ind in c(1:N)){
          # make into matrix
          L_ind <- matrix(0,p,p)
          L_ind[lower.tri(L_ind,diag = F)] <- L_vectors[ind,]
          diag(L_ind) <- rep(1,p)
          Sig_ind <- L_ind %*% D %*% t(L_ind)
          beta_transformed <- beta %*% t(chol(Sig_ind)) # check this
          e_transformed <- e[ind,] %*% t(chol(Sig_ind)) # check this
          X[ind,] <- G[ind,] %*% beta_transformed + e_transformed
      }
      
      
      c <- 2
      Y <- apply(X,1,mdd_risk,threshold=c,n=n)
      prev <- sum(Y)/N
      
      while (prev<0.25){
        c <- c - 0.1
        Y <- apply(X,1,mdd_risk,threshold=c,n=n)
        prev <- sum(Y)/N
      }
      
      print(paste("Set threshold",c,"and prevalence",prev))
      res <- greml(Y,G)
      h2 <- res$h2
      
      rows_hercov2 <- rbind(rows_hercov2,data.frame("h2_symptoms"=her,
                                    "h2_disease"=h2,
                                    "sig2g"=res$sig2g,
                                    "sig2e"=res$sig2e,
                                    "prev"=sum(Y),
                                    "rep"=r,
                                    "P"=p,
                                    "c"=c,
                                    "n"=n,
                                    "hercov"=hercov))
    }
  }
}

rows_hercov2

rows_hercov2 <- rows_hercov2 %>% mutate(h2_l = h2_disease * (prev/N) * (1-prev/N) / dnorm(prev/N)^2)

```



```{r}
ggplot(rows_hercov2,aes(x=hercov,y=h2_disease)) +
  geom_point(aes(col=factor(n+1))) +
  #geom_point(aes(y=h2_l,col="expected h2_l (one symptom)")) +
  labs(x="Covariance heritability",
       col="Number of symptoms past threshold",
       y="Disease heritability")
```


