---
title: "GWAS to find covariance QTLs"
output:
  html_document:
    df_print: paged
---



```{r,include=F}
library(tidyverse)
library(MASS)
library(fastmatrix)
library(tidyverse)
library(matlib)
library(ggcorrplot)

simulate_genotypes <- function(N,L){
  ### function to simulate genotypes
  ps <- rbeta(n = L,4,4)
  G <- matrix(NA,nrow=N,ncol=L)
  for (i in c(1:L)){
    gen <- rbinom(N, size = 2, prob = ps[i])
    G[,i] <- gen
  }
  return(G)
}

give_disease_risk <- function(nodes,f=mean,threshold=5){
  # returns disease risk/severity 
  # risk given by some function of the nodes
  risk <- f(nodes)
  return(risk)
}

# a more funk disease function
mdd_risk <- function(v,threshold=1,n=3){
  # linear combination of node activation - this is just an example
  return(sum(v > threshold)>n)
}

greml <- function( y, G, constrain=F, quantnorm=F ){
  if( quantnorm ){
    ranks <- rank(y)
    quantiles <- qnorm(ranks/(length(ranks) + 1))
    y <- quantiles
  }
  y <- scale(y)
  G <- scale(G)
  N <- nrow(G)
  svdG <- svd(G)
  out <- eigen_lmm( yprime=t(svdG$u) %*% y, Lam.K=(svdG$d/sqrt(N)+1e-12)^2, constrained=constrain )
  sig2g <- out$sig2g
  sig2e <- out$sig2e
  list( h2=sig2g/(sig2g+sig2e), sig2g=sig2g, sig2e=sig2e )
}


eigen_lmm <- function (yprime, Lam.K, constrained = TRUE) {
  if (constrained) {
    delta <-
      optimise(
        eigen_lmm_obj,
        c(0, 1e5),
        maximum = TRUE,
        Lam.K = Lam.K,
        yprime = yprime
      )$maximum
  } else {
    ### these parameters correspond to h^2>0, with h^2>1 for delta<0
    opt_lims_plus <- c(-min(Lam.K), 1e5)
    opt_plus <-
      optimise(
        eigen_lmm_obj,
        opt_lims_plus,
        maximum = TRUE,
        Lam.K = Lam.K,
        yprime = yprime
      )
    ### these parameters correspond to h^2<0
    opt_lims_minus <- c(-1000, -max(Lam.K))
    opt_minus <-
      optimise(
        eigen_lmm_obj,
        opt_lims_minus,
        maximum = TRUE,
        Lam.K = Lam.K,
        yprime = yprime
      )
    delta <-
      ifelse(opt_plus$objective > opt_minus$objective,
             opt_plus$maximum,
             opt_minus$maximum)
  }
  sig2g <- mean(yprime ^ 2 / (Lam.K + delta))
  sig2e <- delta * sig2g
  list(h2 = sig2g / (sig2g + sig2e),
       sig2g = sig2g,
       sig2e = sig2e)
}


eigen_lmm_obj <- function(delta, yprime, Lam.K){
  denominator <- Lam.K + delta
  sig2g <- mean(yprime^2/denominator)
  sum(dnorm( yprime, mean=0, sd=sqrt(sig2g*denominator), log=TRUE ))}

sig_from_row <- function(ind.sig,i,n_node){
  ## function which takes the matrix of individ. level sigmas and builds a covariance matrix which we can draw from
  
  # extract relevant row
  sig_vec <- ind.sig[i,]
  
  # make into matrix
  sig_mat <- matrix(0,n_node,n_node) 
  sig_mat[lower.tri(sig_mat,diag=TRUE)] <- sig_vec
  sig_mat[upper.tri(sig_mat)] <- t(sig_mat)[upper.tri(sig_mat)]
  
  return(sig_mat)
}

scale01 <- function (x)
{
  x <- x - min(x, na.rm = T) + 1e-8
  x <- x/max(x, na.rm = T) - 1e-8
  x
}

theme_set(theme_bw())
```

### Background

Suppose you have a disease, $Y$, which depends on a set of symptoms $X$ through some function $f(X)$. To make this more concrete, we'll use the example of major depressive disorder (MDD), which is diagnosed via a questionnaire asking about a set of symptoms (anxiety, feelings of despair...). The patient muse have at least a certain number of these symptoms to be diagnosed with the disease. 

The symptoms co-vary. Suppose that they co-vary differently in every individual, due to genetics. In fact, someone is more likely to have the disease if their symptoms co-vary more. 

In this notebook, we want to show that a certain amount of heritability can be explained through genetic effects on symptom covariance, even when symptoms themselves are not heritable.

### Model

\begin{align*}
&w   &\sim N(0,h^2_w K + (1-h^2_w) I ) \\
&\Sigma_i &:= w_i J + (1-w_i) I \\
&X_{i,} | \Sigma_i &  N(0,\Sigma_i) \\
&y_i &= f(X_{i,} )
\end{align*}

Where $J$ is a matrix of ones. The higher an individual's covariance weight $w$, the more covariance there is between their symptoms. 

### Simulating genetic effects on the covariance


```{r}
### Defining parameters
N <- 20000 # number of individuals
l <- 100 # number of loci - don't need many for GREML
reps <- 10 # number of replicates
p <- 2 # number of symptoms
hercovs <- seq(0,1,0.1) # heritability of covariance
her <- 0 # heritability of symptoms

# admin
J <- matrix(1,p,p)
rows <- data.frame()

for (r in c(1:reps)){
  print(paste("At replicate",r))
  for (hercov in hercovs){
    # simulate a genotype matrix
    G <- simulate_genotypes(N = N,L = l)
    G <- scale(G)
    # genetic effects on covariance
    beta_cov <- matrix(mvrnorm(n=l,mu = rep(0,1),Sigma = diag(1)),nrow=l) %*% diag(1,1,1) * sqrt(hercov / l)
    e_cov <- matrix(mvrnorm(n=N,mu = rep(0,1),Sigma = diag(1)),nrow=N) %*% diag(1,1,1) * sqrt(1-hercov)
    J_weights <- G %*% beta_cov + e_cov
    J_weights <- scale01(J_weights)
    # genetic effects on symtpoms
    beta <- matrix(mvrnorm(n=l,mu = rep(0,p),Sigma = diag(5,p,p)),nrow=l) %*% diag(1,p,p) * sqrt(her / l)
    e <- matrix(mvrnorm(n=N,mu = rep(0,p),Sigma = diag(5,p,p)),nrow=N) * sqrt(1-her)
    # symptom matrix
    X <- matrix(0,nrow=N,ncol=p)
    # Give each person a covariance matrix and draw symptoms
    for (ind in c(1:N)){
        # get weight
        ind_J_weight <- J_weights[ind]
        Sig_ind <- ind_J_weight * J + (1-ind_J_weight) * diag(1,p,p)
        beta_transformed <- t(chol(Sig_ind)) %*% t(beta) %>% t() # check this
        e_transformed <- t(chol(Sig_ind)) %*% e[ind,] %>% t() # check this
        X[ind,] <- G[ind,] %*% beta_transformed + e_transformed
    }
    # give the disease
    Y <- apply(X,1,mdd_risk,threshold=1,n=1)
    prev <- sum(Y)/N
    # estimating heritabilities
    h2J <- greml(J_weights,G)$h2
    h2IP <- greml(X[,1] * X[,2],G)$h2
    h2Y <- greml(Y,G)$h2
    
    rows <- rbind(rows,data.frame("h2_symptoms"=her,
                              "h2_Y"=h2Y,
                              "h2_IP"=h2IP,
                              "h2_J"=h2J,
                              "prev"=prev,
                              "rep"=r,
                              "P"=p,
                              "c"=c,
                              "n"=n,
                              "hercov"=hercov))
  }
}

rows <- rows %>% mutate(h2_liability = h2_Y * (prev) * (1-prev) / dnorm(prev)^2)

```

### Does the base covariance affect heritability?

```{r}
rows %>% 
  pivot_longer(cols=c("h2_J","h2_liability","h2_IP"),names_to = "feature",values_to = "GREML heritability") %>%
  ggplot(aes(x=hercov,col=feature,y=`GREML heritability`))+
  geom_point() +
  geom_smooth(method="lm",se=F,linetype=2,linewidth=0.5,alpha=0.5) +
  labs(x="true covariance heritability")
```

### How is the covariance heritability affecting disease prevalence?

```{r}
rows %>% 
ggplot(aes(x=hercov,y=prev))+
  geom_point() +
  geom_smooth(method="lm",se=F,linetype=2,linewidth=0.5,alpha=0.5) +
  labs(x="base covariance")
```

Not very much - which is great! Now we inspect the data from the last simulation, where the covariance heritability was 1:

```{r}
phenos <- data.frame(X,Y,J=J_weights) %>%
  mutate(IP12=((X1-mean(X1))* (X2-mean(X2)))/sqrt(var(X1)*var(X2))) # normalised IP
```

```{r}
ggplot(phenos) + geom_point(aes(x=X1,y=X2,col=IP12),shape=16) + scale_color_viridis_c()

ggplot(phenos) + geom_point(aes(x=X1,y=X2,col=Y),shape=16) 

ggplot(phenos) + geom_point(aes(x=J,y=IP12,col=Y)) 

ggplot(phenos) + geom_boxplot(aes(x=J,y=Y)) 
```
And infer effect sizes:

```{r}
apply(G,2,function(x) lm(phenos$IP12 ~ x)$coefficients[2]) -> GWAS_IP
apply(G,2,function(x) lm(phenos$J ~ x)$coefficients[2]) -> GWAS_J
apply(G,2,function(x) lm(phenos$X1 ~ x)$coefficients[2]) -> GWAS_X1
apply(G,2,function(x) lm(phenos$X2 ~ x)$coefficients[2]) -> GWAS_X2
apply(G,2,function(x) lm(phenos$Y ~ x)$coefficients[2]) -> GWAS_Y
data.frame(beta=beta_cov[,1],
           beta_hat_J=GWAS_J,
           beta_hat_IP=GWAS_IP,
           beta_hat_Y=GWAS_Y,
           beta_hat_X1=GWAS_X1,
           beta_hat_X2=GWAS_X2
           ) -> GWAS
```


```{r}
GWAS %>% ggplot(aes(x=beta,y=beta_hat_J)) + geom_point() +geom_smooth(method="lm")
GWAS %>% ggplot(aes(x=beta,y=beta_hat_IP)) + geom_point() +geom_smooth(method="lm")
GWAS %>% ggplot(aes(x=beta,y=beta_hat_X1)) + geom_point() +geom_smooth(method="lm")
GWAS %>% ggplot(aes(x=beta,y=beta_hat_X2)) + geom_point() +geom_smooth(method="lm") 
GWAS %>% ggplot(aes(x=beta,y=beta_hat_Y)) + geom_point() +geom_smooth(method="lm") 
```

The plot of the inferred effect sizes for the disease compared to the true effect sizes is very encouraging!

```{r}
GWAS %>% cor() %>% ggcorrplot(method="circle")
```

